\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2016-04-20 manuscript.tex
updates ahead of lab presentation

\noindent The latent space model (LSM) for network data is a popular model that combine a generalized linear model with a latent spatial embedding of the network to model ties. It has been used to decrease error in the estimation of and inference regarding the effect of observed covariates. In applications of the LSM, it is assumed that the latent spatial embedding can control for unmeasured confounders that have an effect on the value of edges in a network. We investigate the latent space model's performance on this task via a Monte Carlo simulation wherein we show that in the presence of an unmeasured covariate which can be appropriately modeled using a latent space, estimation and inferential error does decrease. However, when no such unmeasured covariate is present, a generalized linear model substantially outperforms the latent space model. The LSM is often biased downwards by nearly half the magnitude of the coefficient of the observed covariate. This pattern is repeated with generalization error and with type 1 and type 2 error rates. We investigate solutions to this issue and conclude with recommendations as to the appropriate use of the latent space model for observational studies.

@@

Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models. The latent space model (LSM) \cite{hoff2002latent} for network data is regularly used to estimate the effect of covariates in the presence of latent network structure. However, it is unclear that the introduction of a latent space decreases the expected error for the parameter(s) of interest when aspects of the network structure (e.g., homophily) that are unmeasured are correlated with measured covariates. There are two reasons that using the LSM may lead to increased error. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spurriously correlated with the observed covariates, leading to reduced efficiency due to multicollinearity. Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable, while the latent space parameters are used to model other sources of variation.

@@

\citep{mackinnon1989collinearity} and \citep{weissfeld1991multicollinearity} investigate multicollinearity in generalized linear models and propose diagnostic criteria. Collinearity of the covariates causes the observed information matrix to be ill-conditioned, which produces nmerical instability in the iterative esimation of $\hat{\mathbf{\beta}}$. \citep{weissfeld1991multicollinearity}, extending the work of \citep{davis1986example}, propose the use of the condition index and the variance decomposition proportion of each condition index, to diagnose the presence and degree of multicollinearity in generalized linear models. The observed information matrix at the maximum likelihood estimate, $\mathcal{I}(\hat{\mathbf{\beta}})$ is scaled to have unit length and singular value decomposition is applied. The resulting right singular values are the eigenvectors and the associated singular values are the square root of the eigenvalues of the scaled observed information matrix since it is real Hermitian. The condition index is the ratio of the largest eigenvalue to each of the eigenvalues, and is at least 1. The square root of the condition number gives the potential magnification of the numerical error in inverting the observed information matrix. Hence our expectation is that as collinearity increases the variance of estimators of the effect of observed covariates will increase regardless of whether this is ommitted confounding structure in the data.

@@

If we assume that the $z$ are independent and distributed $N(0,\sigma^2)$, it is straightforward to derive the distribution of $ |z_i - z_j|$.\footnote{The assumption that $z \sim N(0,\sigma^2$) is consistent with $z$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim N(0,2\sigma^2)$ \citep{devore2012}. The normality of $z_i - z_j$ implies that $|z_i - z_j|$ has a folded normal distribution \citep{leone1961}. Following from this result, we know that the variance of  $|z_i - z_j|$ is $$\sigma^2(2-4/\pi). $$ Let $\alpha^2$ be the empirical variance of $x$. Then the variance in the linear predictor is $\beta_1^2\alpha^2$. Given an estimate $\hat{\beta_1}$, we can set the variance of the latent space prior to approximately $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$ in order to tune the prior to avoid replicating $x$ via the distances between latent coordinates. This result is limited in that it requires the assumption of normally distributed latent coordinates and applies only to a single latent dimension. However, it may provide a reasonable approximation in the case of non-normal coordinates and/or multiple dimensions.

\cite{thirey2015} derive the distribution of Euclidean Distances between two $k$-dimensional points where each coordinate is drawn from a standard normal distribution.

\subsection{Tasks}
\begin{itemize}
\item Review multicollinearity issues in GLM (ZJ)
\item Describe how multicollinearity will manifest in LSM (ZJ)
\item Delve into LSM priors (BD)
\begin{itemize}
\item Scaling variance to observed covariates
\item Spike and slab
\item Covariance
\end{itemize}
\end{itemize}

@@

To study how the LSM behaves in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space. For simplicity we consider a unidimensional latent space. To generate an observed covariate which has a controllable collinearity with the latent network structure we follow a three-step process. First, we simulate $k$-dimensional positions for each node, drawing each from a normal distribution, and calculate dyadic distances $d$ between each pair of positions. Second, given a target covariance matrix ($\Sigma$) among the covariates and distances, $\langle X,d \rangle$, we derive the conditional mean vector and covariance, assuming that $X$ has a joint normal distribution given $d$ (see \cite[pp. 116--117]{eaton1983} for the conditional normal derivation). Third, we simulate $X$ as a multivariate normal random variable with the respective conditional means and covariance. Finally we standardize each $X$ to have 0 mean and unit variance. To generate the covariance matrix $\Sigma$ which controls the dependence between the omitted network structure and the observed covariate $X$ we utilize the C-vine method of \cite{lewandowski2009generating}. % fill in a basic description of the method here

We consider three exponential family distributions for the edges: Gaussian, Binomial, and Poisson. Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space $k$ at 1 for simplicity, and consider 5 values of the collinearity parameter $\eta$, equally spaced between $.75$ and $100$, where $.75$ corresponds to near perfect collinearity and $100$ to independence.

We standardize $X$ to have a variance of $1$ and a mean of $0$, and the linear predictor, $X \beta - d$

\begin{figure}
\includegraphics[width=\textwidth]{max_r_vine.png}
\caption{The distribution of the absolute value of the maximum off-diagonal correlation.}
\end{figure}

Each step in the simulation involves the estimation of a GLM, the LSM, and the LSM with a prior on the latent space which is scaled by $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$, where $\alpha^2$ is the empirical variance of $X$. At each step in the simulation, the LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R}. An initial run of 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler. Every 100th iteration is saved. Convergence in the log probability of the model is assessed using the Geweke diagnostic. If the convergence criteria is satisfied the simulation continues to the next set of arguments, otherwise the number of iterations is doubled. If the convergence criteria is still not satisfied, then the aforementioned step in the simulation is flagged for review. At each point in the simulation's parameter space, we execute 500 Monte Carlo iterations.

We first compute the MLE estimated by iteratively reweighted least squares in \texttt{R}. For the LSM we use the posterior mode. In the cases where edges are Binomial or Poisson distributed, we scale the estiamtes % figure out how to do this.

To evaluate estimation error we compute the bias: the difference between the true coefficient $\beta_1$ and the estimate. Figure \ref{fig:estimation} shows the results. When a latent space is present (i.e., $\beta_2 = -1$) the LSM outperforms the GLM in terms of bias across the board. However, when latent network structure is not present ($\beta_2 = 0$) the LSM is sharply biased downwards, indicating that the latent space is representing variability in the edges that is attributable to the observed covariate $X$.

\begin{figure}
\includegraphics[width=\textwidth]{estimation.png}
\caption{The bias of estimates of the effect of the observed covariate $X$. The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The $y$-axis gives a Monte Carlo estimate of the bias, and the grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate. The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels. Each panel represents 3 values of $\eta$ with 500 Monte Carlo iterations executed at each point.
\label{fig:estimation}}
\end{figure}

We consider networks of varying size (i.e. number of nodes), density (i.e., the number/strength of edges between nodes), and collinearity structure between the ommitted distance between nodes and the measured covariates. For each point in this space we evaluate the mean square prediction error of the model on new data drawn from the same distribution (i.e., the generalization error), the bias of the estimated coefficients for the measured covariate, the type 2 error rate (i.e., the probability that the null hypothesis that each regression parameter is $0$ is incorrectly accepted), and the type 1 error rate, that is, the probability of rejecting the null hypothesis that the effect of the measured covariate is 0 when it is in fact 0. Inference regarding $\boldsymbol\beta$ is computed using the maximum a posteriori estimate (i.e., the mode of the posterior), and credible sets for $\boldsymbol\beta$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution for each regression parameter.

\begin{figure}
\includegraphics[width=\textwidth]{inference.png}
\caption{Monte Carlo estimates of the Type-1 and Type-2 error regarding the effect of the observed covariate $X$ are shown on the $y-axis$. When the right gray panel indicates ``Type-1'' $\beta_1 = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta_1 = 0$ is falsely rejected. When the right gray panel indicates ``Type-2'' then $\beta_1 = 1$ and the error rate shown in each panel in that row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis. The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate. The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels. Each panel represents 5 values of $\eta$ with 500 Monte Carlo iterations executed at each point. \label{fig:inference}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{generalization.png}
\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the generative model using a model trained on independent data. The $x$-axis gives the value of the parameter $\eta$, which controls the degree of dependence between $X$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The grey ribbon gives a pointwise 95\% confidence interval for the Monte Carlo estimate. The number of nodes and the presence or absence of omitted network structure are indicated in the top panels, while the distributional family for the edges and the value of the coefficient for $X$, $\beta_1$ are indicated on the right panels. Each panel represents 5 values of $\eta$ with 500 Monte Carlo iterations executed at each point. \label{fig:generalization}}
\end{figure}

@@

>>>>

\noindent The latent space model (LSM) for network data is a popular model that combine a generalized linear model with a latent spatial embedding of the network to model ties. It has been used to decrease error in the estimation of and inference regarding the effect of observed covariates. In applications of the LSM, it is assumed that the latent spatial embedding can control for unmeasured confounders that have an effect on the value of edges in a network. We investigate the latent space model's performance on this task via a Monte Carlo simulation wherein we show that in the presence of an unmeasured covariate which can be appropriately modeled using a latent space, estimation and inferential error decreases, however, the difference is not large, even at high levels of multicollinearity between the unmeasured and observed covariates. However, the prediction error of the LSM when an unmeasure covariate is present is substantially lower in most cases. When an omitted covariate is not present, the LSM can introduce bias and increased inferential error above that of a generalized linear model. We investigate solutions to this issue and conclude with recommendations as to the appropriate use of the latent space model for observational studies.

@@

Latent variable inference, generally conceived, present the possibility of representing unmeasured data in statistical models. The latent space model (LSM) \cite{hoff2002latent} for network data is regularly used to estimate the effect of covariates in the presence of latent network structure. However, it is unclear that the introduction of a latent space decreases the expected error for the parameter(s) of interest when aspects of the network structure (e.g., homophily) that are unmeasured are correlated with measured covariates. There are two reasons that using the LSM may lead to increased error. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spurriously correlated with the observed covariates, leading to reduced efficiency due to multicollinearity. Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable, while the latent space parameters are used to model other sources of variation.

@@

\citep{mackinnon1989collinearity} and \citep{weissfeld1991multicollinearity} investigate multicollinearity in generalized linear models and propose diagnostic criteria. Collinearity of the covariates causes the observed information matrix to be ill-conditioned, which produces numerical instability in the iterative estimation of $\hat{\mathbf{\beta}}$, the ``effects'' of covariates included in a regression. \citep{weissfeld1991multicollinearity}, extending the work of \citep{davis1986example}, propose the use of the condition index and the variance decomposition proportion of each condition index, to diagnose the presence and degree of multicollinearity in generalized linear models. The observed information matrix at the maximum likelihood estimate, $\mathcal{I}(\hat{\mathbf{\beta}})$ is scaled to have unit length and singular value decomposition is applied. The resulting right singular values are the eigenvectors and the associated singular values are the square root of the eigenvalues of the scaled observed information matrix since it is real Hermitian. The condition index is the ratio of the largest eigenvalue to each of the eigenvalues, and is at least 1. The square root of the condition number gives the potential magnification of the numerical error in inverting the observed information matrix. Hence our expectation is that as collinearity increases the variance of estimators of the effect of observed covariates will increase regardless of whether this is ommitted confounding structure in the data.

@@

If we assume that the $z$ are independent and distributed $N(0,\sigma^2)$, it is straightforward to derive the distribution of $ |z_i - z_j|$.\footnote{The assumption that $z \sim N(0,\sigma^2$) is consistent with $z$ being drawn from the prior that is conventionally used in the LSM.} Since it is a special case of a ``normal difference distribution'', we know that $z_i - z_j \sim N(0,2\sigma^2)$ \citep{devore2012}. The normality of $z_i - z_j$ implies that $|z_i - z_j|$ has a folded normal distribution \citep{leone1961}. Following from this result, we know that the variance of  $|z_i - z_j|$ is $$\sigma^2(2-4/\pi). $$ Let $\alpha^2$ be the empirical variance of $x$. Then the variance in the linear predictor is $\beta_1^2\alpha^2$. Given an estimate $\hat{\beta_1}$, we can set the variance of the latent space prior to approximately $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$ in order to tune the prior to avoid replicating $x$ via the distances between latent coordinates. This result is limited in that it requires the assumption of normally distributed latent coordinates and applies only to a single latent dimension. However, it may provide a reasonable approximation in the case of non-normal coordinates and/or multiple dimensions. \cite{thirey2015} derive the distribution of Euclidean Distances between two $k$-dimensional points where each coordinate is drawn from a standard normal distribution.

@@

To study how the LSM behaves in the situation in which we have some observed covariates as well as omitted network structure which can be represented using a Euclidean latent space. For simplicity we consider a unidimensional latent space. To generate an observed covariate which has a controllable collinearity with the latent network structure we follow a three-step process. First, we simulate $k$-dimensional positions for each node, drawing each from a normal distribution, and calculate dyadic distances $d$ between each pair of positions. Second, given a target covariance matrix ($\Sigma$) among the covariates and distances, $\langle X,d \rangle$, we derive the conditional mean vector and covariance, assuming that $X$ has a joint normal distribution given $d$ (see \cite[pp. 116--117]{eaton1983} for the conditional normal derivation). Third, we simulate $X$ as a multivariate normal random variable with the respective conditional means and covariance. Finally we standardize each $X$ to have zero mean and unit variance. To generate the covariance matrix $\Sigma$ which controls the dependence between the omitted network structure and the observed covariate $X$ we utilize the C-vine method of \cite{lewandowski2009generating}. % fill in a basic description of the method here

We consider three exponential family distributions for the edges: Gaussian, Binomial, and Poisson. Additionally we control the number of nodes in the network $n = 25, 50, 100$, fix the dimension of the latent space $k$ at 1 for simplicity, and consider 4 values of the collinearity parameter $\eta$, $.1$, $1$, $100$, and $1000000$, where $.1$ a corresponds to near perfect collinearity and $1000000$ to independence. See Figure \ref{fig:vine} for the distribution of the absolute value of the maximal off-diagnoal correlation generated at each value of $\eta$.

\begin{figure}
\includegraphics[width=\textwidth]{max_r_vine.png}
\caption{The distribution of the absolute value of the maximum off-diagonal correlation. \label{fig:vine}}
\end{figure}

Each step in the simulation involves the estimation of a GLM, and the LSM. We consider the LSM with several different priors on the coefficient for the observed covariate and the latent space. We set the prior variance of $\beta_1$ to be either $1$ or $10$ and alternatively use a diffuse normal prior on the latent space or scale it by $\hat{\beta_1}^2\alpha^2/(2-4/\pi)$, where $\alpha^2$ is the empirical variance of $X$.

At each step in the simulation, the LSM is estimated using the canonical implementation in the \texttt{latentnet} package in \texttt{R}. An initial run of 10,000 burn in iterations, followed by 1,000,000 iterations of the sampler. Every 100th iteration is saved. Convergence in the log probability of the model is assessed using the Geweke diagnostic. If the convergence criteria is satisfied the simulation continues to the next set of arguments, otherwise the number of iterations is doubled. If the convergence criteria is still not satisfied, then the aforementioned step in the simulation is flagged for review. At each point in the simulation's parameter space, we execute 500 Monte Carlo iterations.

@@

We first compute the MLE estimated by iteratively reweighted least squares in \texttt{R}. For the LSM we use the posterior mode. In the cases where edges are Binomial distributed, we scale the estiamtes using the reciprocal of the bias of $\beta_1$ when $X$ and $d$ are independent: $\frac{\sqrt{3.28 + \beta_2^2 \text{Var(d)}}}{\sqrt{3.29}}$, where $3.29$ is the variance of a standard logistic distribution (See the derivation of the bias under an independent but ommitted covariante in \cite{mood2010logistic}).

We consider networks of varying size (i.e. number of nodes) and collinearity structure between the ommitted distance between nodes and the measured covariates. For each point in this space we evaluate the mean square prediction error of the model on new data drawn from the same distribution (i.e., the generalization error), the bias of the estimated coefficients for the measured covariate, the type 2 error rate (i.e., the probability that the null hypothesis that each regression parameter is $0$ is incorrectly accepted), and the type 1 error rate, that is, the probability of rejecting the null hypothesis that the effect of the measured covariate is 0 when it is in fact 0. Inference regarding $\beta_1$ is computed using the maximum a posteriori estimate (i.e., the mode of the posterior), and credible intervals for $\beta_1$ are defined by using the region of highest posterior density which covers $95\%$ of the marginal posterior distribution for each regression parameter.

To evaluate estimation error we compute the bias: the difference between the true coefficient $\beta_1$ and the estimate. Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results. When a latent space is present (i.e., $\beta_2 = -1$, shown in Figure \ref{fig:estimation_ls}) the LSM outperforms the GLM in terms of bias across the board. However, when latent network structure is not present ($\beta_2 = 0$, shown in Figure \ref{fig:estimation_nls}) the LSM is biased upwards in the Binomial case and appears to have higher variance (perhaps due to efficiency losses) for Poisson and Gaussian edges (preliminary interpretation).

\begin{figure}
\includegraphics[width=\textwidth]{estimation_ls.png}
\caption{The bias of estimates of the effect of the observed covariate $X$ when there is an ommitted variable. The $x$-axis gives the value of the parameter $\eta$ which controls the degree of dependence between $X$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The $y$-axis gives a Monte Carlo estimate of the bias. The number of nodes are indicated in the top panels, while the distributional family of the edges is shown on the right panel. Each panel represents 4 values of $\eta$ with 500 (partially complete) Monte Carlo iterations executed at each point.
\label{fig:estimation_ls}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{estimation_nls.png}
\caption{The bias of estimates of the effect of the observed covariate $X$ when there is no ommitted variable.
\label{fig:estimation_nls}}
\end{figure}

Figures \ref{fig:inference_type_1} and \ref{fig:inference_type_2} show the inferential error rates of the LSM with different priors for $\beta_1$ and the latent space compared to a GLM. Type 1 error rates for the LSM are in general lower than those of a GLM when an unmeasured covariate exists, however, when one does not exist the GLM outperforms the LSM. However, scaling the prior of the latent space to match that of the observed covariates appears to make the LSM's error rate comparable to that of the GLM. Type 2 error rates are very similar for the GLM and the LSM with all evaluated prior settings; to the point that they are indistinguishable.

\begin{figure}
\includegraphics[width=\textwidth]{inference_type_1.png}
\caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $X$ are shown on the $y$$-axis. Here, $\beta_1 = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta_1 = 0$ is falsely rejected. \label{fig:inference_type_1}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{inference_type_2.png}
\caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $X$ are shown on the $y-axis$. Here, $\beta_1 = 1$, and the error rate shown in each panel in each row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis. \label{fig:inference_type_2}}
\end{figure}

Generalization error, in this case defined as the expected error on new edges generated using a fixed set of latent distances between the nodes and a node-level covariate is shown in figure \ref{fig:generalization}. In nearly all cases the LSM substantially outperforms the GLM, especially when the ommitted covariate is not highly collinear with the observed covariate.

\begin{figure}
\includegraphics[width=\textwidth]{generalization.png}
\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the generative model using a model trained on independent data. \label{fig:generalization}}
\end{figure}

\section{Conclusion}

Although the results are (very) preliminary (and have been plagued by my incompetence!), it seems increasingly clear that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space to uncover patterns of interest or to predict edges in networks where there is likely to be omitted structure that is not adequately modelled using measured covariates. However, inferential and estimation error are often smaller than that of a GLM when an omitted variable that corresponds to a form that can be modelled using a latent space is present, and can lead to an up to $25\%$ decrease in bias relative to a GLM. Whether or not there is a reduction in estimation and/or inferential error depends on the strength of the ommitted confounder, sample size (the LSM having many more parameters than an equivalent GLM), and relies on the ommitted confounder being representable as a latent space. Lastly, there is some (preliminary) evidence to suggest that scaling the prior of the latent space to match that of the measured covariates can reduce Type-1 error.
